---
title: 'Assignment 4: Exploratory report on juvenile snowshoe hares in Bonanza Creek
  Experimental Forest'
author: "Mia Guarnieri"
date: "11/7/2021"
output: 
  html_document: 
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(lubridate)
library(dplyr)
library(effsize)
library(kableExtra)
```

#Introduction

A 4 - 5 sentence "mini-introduction" to what this study is exploring, including basic background information about the subject, site and study (with citations as needed), and a brief overview of what is included in the exploratory report. You should probably do this last so you know what’s contained in the report.

#Data and analyses

A 3 - 4 sentence section in which you briefly describe the data, including citation & credit as needed, and provide an overview of the main analysis, statistical methods and tools used.

#Juvenile hare initial investigation

##Annual juvenile hare trap counts 

Count the total number of juvenile hare trappings during each year of the study, and create a finalized data visualization of the counts by year. Include a figure caption below the figure in your report. 

```{r}
#read in the hare data
bonanza_hares <- read.csv(here("Data", "bonanza_hares.csv"))



```

